International Relations are a Social Network:\\ A Latent Network Approach for Global Politics

### 2017-10-17 paper/2_model.tex
sim stuff

To uncover the underlying structure that relational data comes with it is helpful to shift towards restructuring dyadic data in the form of a matrix -- often called an adjacency matrix -- as shown in Figure~\ref{fig:adjMatDeps}. Rows designate the senders of an event and columns the receivers. In the context of events, the cross-sections in this matrix represent the actions that were sent by an actor in the row to those designated in the columns. Thus $y_{ij}$ designates an action $y$, such as a conflictual event or trade flows, that is sent from actor $i$ to actor $j$.

Using the structure of an adjacency matrix, we can visualize the types of first- and second-order dependencies that complicate the analysis of relational data in traditional GLMs. Figure~\ref{fig:adjMatDeps} clarifies the types of dependencies that can manifest in these types of data structures. The adjacency matrix on the top left highlights a particular row of an adjacency matrix, to illustrate that these values may be more similar to each other than other values in the adjacency matrix because each has a common sender. Homogeneity in interactions involving a common sender also manifest heterogeneity in how active actors are across the network when compared to each other. Thus in most relational datasets (e.g., trade flows, conflict, participation in international organizations, even networks derived from Twitter or Facebook) we often find that there are some actors which are much more active than others \citep{barabasi:reka:1999}. Unless one is able to develop a model that can account for the variety of explanations that may play a role in determining why a particular actor may be more active than others, parameter estimates from standard statistical models will be biased.

@@

The LFM contribution to the AME comes in the multiplicative term: $\textbf{u}_{i}^{\top} \textbf{D} \textbf{v}_{j}=\sum_{k \in K} d_{k} u_{ik} v_{jk}$. This model posits a latent vector of characteristics $\mathbf{u_{i}}$ and $\mathbf{v_{j}}$ for each sender $i$ and receiver $j$. The similarity or dissimilarity of these vectors will then influence the likelihood of activity, and provides a representation of third order interdependencies \citep{minhas:etal:2016:arxiv}. The representation of third order interdepedencies is accomplished by a process similar to computing the singular value decomposition (SVD) of the observed network. When taking the SVD we factorize our observed network into the product of three matrices: $\textbf{U}, \textbf{D}, \text{ and }, \textbf{V}$. This provides us with a low-dimensional representation of our original network in terms of third order dependence patterns arising in the network.\footnote{The dimensions of $\textbf{U}$ and $\textbf{V}$ are $n \times K$ and $\textbf{D}$ is a $K \times K$ diagonal matrix.} Specifically, values in $\textbf{U}$ provide a representation of how stochastically equivalent actors are as senders in a network, or more simply put how similar actors are in terms of who they are initiating, for example, conflict with. $\hat{\textbf{u}}_{i} \approx \hat{\textbf{u}}_{j}$ would indicate that actor $i$ and $j$ initiate battles with similar third actors. $\textbf{V}$ provide a similar representation but from the perspective of how similar actors are as receivers. The values in $\textbf{D}$, a diagonal matrix, represent levels of homophily in the network.\footnote{Unlike traditional SVD, in the latent factor model, the singular values are not restricted to be positive, thus allowing us to account for both the presence and absence of homophily.} Parameter estimation in the AME takes place within the context of a Gibbs sampler in which we iteratively sample from the posterior distribution of the full conditionals for each parameter.

 Specifically, given initial values of $\{\bm\beta, \textbf{a}, \textbf{b}, \textbf{U}, \textbf{V}, \Sigma_{ab}, \rho, \text{ and } \sigma_{\epsilon}^{2}\}$, the algorithm proceeds as follows until convergence:

>>>>

To uncover the underlying structure that relational data comes with it is helpful to shift towards restructuring dyadic data in the form of a matrix -- often referred to as an adjacency matrix -- as shown in Figure~\ref{fig:adjMatDeps}. Rows designate the senders of an event and columns the receivers. In the context of events, the cross-sections in this matrix represent the actions that were sent by an actor in the row to those designated in the columns. Thus $y_{ij}$ designates an action $y$, such as a conflictual event or trade flows, that is sent from actor $i$ to actor $j$.

Using the structure of an adjacency matrix, we visualize in Figure~\ref{fig:adjMatDeps} the types of first- and second-order dependencies that complicate the analysis of relational data in traditional GLMs. The adjacency matrix on the top left highlights a particular row of an adjacency matrix, to illustrate that these values may be more similar to each other than other values in the adjacency matrix because each has a common sender. Homogeneity in interactions involving a common sender also manifest heterogeneity in how active actors are across the network when compared to each other. Thus in most relational datasets (e.g., trade flows, conflict, participation in international organizations, even networks derived from Twitter or Facebook) we often find that there are some actors which are much more active than others \citep{barabasi:reka:1999}. Unless one is able to develop a model that can account for the variety of explanations that may play a role in determining why a particular actor may be more active than others, parameter estimates from standard statistical models will be biased.

@@

The LFM contribution to the AME comes in the multiplicative term: $\textbf{u}_{i}^{\top} \textbf{D} \textbf{v}_{j}=\sum_{k \in K} d_{k} u_{ik} v_{jk}$. This model posits a latent vector of characteristics $\mathbf{u_{i}}$ and $\mathbf{v_{j}}$ for each sender $i$ and receiver $j$. The similarity or dissimilarity of these vectors will then influence the likelihood of activity, and provides a representation of third order interdependencies \citep{minhas:etal:2016:arxiv}. The representation of third order interdepedencies is accomplished by a process similar to computing the singular value decomposition (SVD) of the observed network. When taking the SVD we factorize our observed network into the product of three matrices: $\textbf{U}, \textbf{D}, \text{ and }, \textbf{V}$. This provides us with a low-dimensional representation of our original network in terms of third order dependence patterns arising in the network.\footnote{The dimensions of $\textbf{U}$ and $\textbf{V}$ are $n \times K$ and $\textbf{D}$ is a $K \times K$ diagonal matrix.} Specifically, values in $\textbf{U}$ provide a representation of how stochastically equivalent actors are as senders in a network, or more simply put how similar actors are in terms of who they are initiating, for example, conflict with. $\hat{\textbf{u}}_{i} \approx \hat{\textbf{u}}_{j}$ would indicate that actor $i$ and $j$ initiate battles with similar third actors. $\textbf{V}$ provide a similar representation but from the perspective of how similar actors are as receivers. The values in $\textbf{D}$, a diagonal matrix, represent levels of homophily in the network.\footnote{Unlike traditional SVD, in the latent factor model, the singular values are not restricted to be positive, thus allowing us to account for both the presence and absence of homophily.}

Parameter estimation in the AME takes place within the context of a Gibbs sampler in which we iteratively sample from the posterior distribution of the full conditionals for each parameter. Specifically, given initial values of $\{\bm\beta, \textbf{a}, \textbf{b}, \textbf{U}, \textbf{V}, \Sigma_{ab}, \rho, \text{ and } \sigma_{\epsilon}^{2}\}$, the algorithm proceeds as follows until convergence:

### 2017-10-17 paper/3_sim.tex
sim stuff

Here we consider inference for a regression parameter $\beta$ of a linear or generalized linear model for a network in the case where there is an omitted variable. The true data-generating models we consider are of the form

$y_{i,j} \sim  \mu + \beta x_i x_j + \gamma w_i w_j + \epsilon_{i,j} $

where $Y= \{y_{i,j}\}\in \mathbb R^{n\times n}$ is an observed sociomatix, $x = \{x_i \} \in \mathbb R^n$ is a vector of observed node-specific characteristics, and $w = \{ w_i\} \in \mathbb R^n$ is a vector of unobserved node-specific characteristics. We compare inference for $\beta$ using three models:

\begin{itemize}
	\item A naive regression model assuming independent errors; 
	\item A latent factor model; 
	\item An "oracle" regression model that includes both $x_{i,j}=x_i x_j$ and $w_{i,j}=w_i w_j$. 
\end{itemize}

We make these comparisons in the context of a binary network outcome with a probit regression model.

>>>>

We utilize a simulation study to highlight the utility of AME as an inferential tool for dyadic analysis. We are particularly interested in whether AME can provide unbiased and well calibrated estimates of coefficient parameters in the presence of unobserved network dependencies. Specifically, we are concerned with conducting inference on a regression parameter $\beta$ of a linear model for a network in the case where there is an omitted variable. Say that the true data-generating process for a particular $Y$ is given by:

$y_{i,j} \sim  \mu + \beta x_{i,j} + \gamma w_{i,j} + \epsilon_{i,j} $

where $Y= \{y_{i,j}\}\in \mathbb R^{n\times n}$ is an observed sociomatix, $x = \{x_{i,j} \} \in \mathbb R^{n \times n}$ is a matrix of observed dyad-specific characteristics, and $w = \{ w_{i,j}\} \in \mathbb R^{n \times n}$ is a matrix of unobserved dyad-specific characteristics. We compare inference for $\beta$ using three models:

\begin{itemize}
	\item A naive regression model assuming independent errors; 
	\item A latent factor model; 
	\item An ``oracle'' regression model that includes both $x_{i,j}=x_i x_j$ and $w_{i,j}=w_i w_j$. 
\end{itemize}

We make these comparisons in the context of a linear network outcome with the AME probit regression model described in the previous section.
